mathml: Translate R expressions to MathML and LaTeX/MathJax

### 2022-12-29 Draft3.Rmd
update christmas

---
title: "mathml: Translate R expressions to MathML and LaTeX/MathJax"
date: "2022-12-09"
abstract: >
  This R\ package translates R\ objects to suitable entities in MathML or LaTeX,
  thereby enabling pretty mathematical representation of R\ functions or
  algorithms in data analysis, scientific reports and interactive web content.
  R\ code text descriptions and mathematical content already coexist,
  side-by-side, in the RMarkdown document rendering language. The present
  package allows to use the same R\ objects for both data analysis and
  typesetting in documents or web content. By tightening the link between the
  statistical analysis and its verbal or symbolic representation, this new
  feature is a futher step towards reproducible science. User-defined hooks
  allow to extend the package by mapping new variable names and functions to
  new MathML and LaTeX entities. We provide a few working examples that
  illustrate the use of the package in a little scientific report and in an
  interactive web page.
draft: true
author:  
  # see ?rjournal_article for more information
  - name: Matthias Gondan
    affiliation: University of Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/gondan-rochon/index.html.en
    orcid: 0000-0001-9974-0057
    email:  Matthias.Gondan-Rochon@uibk.ac.at
  - name: Irene Alfarone
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/alfarone/index.html.en
    email: Irene.Alfarone@uibk.ac.at
    orcid: 0000-0002-8409-8900
    affiliation: University of Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib
preamble:
- \usepackage{longtable}
- \usepackage{cancel}
---

@@

The R\ extension of the markdown language [@Xie2020] enables reproducible
statistical reports with nice typesetting in HTML, Microsoft Word, and Latex.
Since recently [@R, version 4.2], R\'s manual pages include support for
mathematical expressions [@Sarkar2022], which already is a big improvement.
However, rules for the translation of R\'s built-in language elements to their
mathematical representation are still lacking. So far, R\ expressions such
as `pbinom(k, N, p)` are printed as they are; pretty mathematical formulae
like $P_\mathrm{Bi}(X \le k; N, p)$ require explicit Latex commands, that
is, `P_\mathrm{Bi}\left(X \le k; N, p\right)`. Except for minimalistic examples,
these commands are tedious to type in and hard to read in their source code.

The present R\ package defines a set of rules for an _automatic_ translation of
R\ expressions to mathematical output in RMarkdown documents [@Xie2020] and
ShinyApp webpages [@Chang2022]. The translation is done by an embedded Prolog
interpreter that maps nested expressions recursively to MathML and
LaTeX/MathJax, respectively. User-defined hooks enable to extend the set of
rules, for example, to represent specific R\ elements by custom mathematical
signs.

The main feature of the package that the same R\ expressions and equations are
used for both mathematical typesetting and calculations. This saves time and
reduces mistakes, as well be illustrated below.

The paper is organized as follows. Section\ 2 presents the background of the
package, including Prolog and the R\ package \pkg{rolog} and the two main classes
of rules for translating R\ objects to mathematical expressions. Section\ 3
illustrates the main features of the \pkg{mathml} package, potential issues and
workarounds using examples from the day-to-day perspective of the user.
Sections\ 4 and\ 5 present two case studies for the use of the package in a
scientific report and a ShinyApp web page. Section\ 6 concludes with a
discussion and ideas for further development. A complete list of the R\ elements
covered by the package is found in the Appendix.

@@

The translation of R\ expressions to mathematical output is achieved through a
Prolog interpreter. Prolog is a classical logic programming language with many
applications in expert systems, computer linguistics and symbolic artificial
intelligence. The main strength of Prolog is its concise representation of facts
and rules for the representation of knowledge and grammar, as well as its
efficient built-in search engine for closed world domains. As it is well-known,
R\ is a statistical programming language for data analysis and statistical
modeling which is widely used in academia and industry. Besides the core
library, a lot of packages have been developed for all kinds of statistical
problems, including statistics-based artificial intelligence tools such as
neural networks for machine learning and deep learning. Whereas Prolog is weak
in statistical computation, but strong in symbolic manipulation, the converse
may be said for the R\ language. The \pkg{rolog} package bridges this gap by
providing an interface to a SWI-Prolog distribution in R. The communication
between the two systems is mainly in the form of queries from R\ to Prolog, but
two predicates allow Prolog to ring back and evaluate terms in R.

For a first illustration of the package, we consider the binomial probability.

@@

The term is quoted to avoid its immediate evaluation which would raise an error
anyway since the variables `k`, `N`, `p` have not yet been defined. Experienced
readers will remember that the quoted expression is a short form for

@@

As is seen from the output above, the variable `term` is not assigned the result
of the calculation, but an R\ "call"
[see, e.g., @Wickham2019, for details on "non-standard evaluation"]. This call
can eventually be evaluated with `eval(...)`,

@@

The R\ package \pkg{mathml} can now be used to render the call in MathML, that
is the math dialect for HTML webpages or in MathJax/LaTeX, as shown
below (some of the curly braces are not really needed in this simple example,
but are necessary in edge cases).

@@

We can include the output in a RMarkdown document by specifying `results='asis'`
in the R\code chunk, as is done in the next example. The
R\ function `mathout(...)` is just a wrapper that invokes `mathml(...)` for HTML
output and `mathjax(...)` for LaTeX output.

@@

At the Prolog end, there is a predicate `math/2` that translates the
call `pbinom(K, N, Pi)` into a "function" `fn/2` with the name `P_Bi`, one
argument `X =< K`, and the two parameters `N` and `Pi`.

@@

Thus, the predicate `math/2` could be considered a "macro" that translates a
mathematical element (here, `pbinom(K, N, Pi)`) to a different mathematical
element, namely `fn(..., ...)`. The low-level predicates `ml/3` and `jax/3`
convert mathematical elements to MathML or LaTeX, respectively.

```prolog
ml(Flags, fn(Name, (Args ; Params)), M)
 => ml(Flags, Name, N),
    ml(Flags, paren(list(op(;), [list(op(','), Args), list(op(','), Params)])), X),
    M = mrow([N, mo(&(af)), X]).
```

The relevant rule for `ml/3` builds the MathML entity `mrow([N, mo(&(af)), X])`,
with `N` representing the function's name and `X` its arguments and parameters,
enclosed in parentheses. The flags can be used for context-sensitive
translation (see, e.g., the errors below).

@@

\pkg{mathml} contains several basic elements used in everyday mathematical
expressions, such as numbers, Latin and Greek letters, multi-letter identifiers,
accents, subscripts, and superscripts.

```{r, results='asis'}
term <- quote(1 + 2L + "a" + a + abc + phi + Phi + varphi + hat(b)[i, j]^2L)
mathout(term)

@@

@@

Basic arithmetic operators and parentheses are translated as they are, as
illustrated below.

```{r, results="asis"}
term <- quote(a - ((b + c)) - d*e + f*(g + h) + i/j + k^(l + m) + (n*o)^{p + q})
mathout(term)
```

For multiplications involving only numbers and symbols, the multiplication sign
is omitted. This heuristic does not always produce the desired result;
therefore, \pkg{mathml} defines the extra functions `dot/2`, `nodot/2`,
and `times/2` that produce the respective multiplication signs.

```{r, results="asis"}
term <- quote(dot(d, e) + nodot(f, g + h) + times(X, Y))
mathout(term)
```

For the standard operators with known precedence, \pkg{mathml} is generally able
to detect if parentheses are needed; for example, parentheses are automatically
put around `g + h` in the `nodot`-example. However, we note unecessary
parentheses around `l + m` above. This is parenthesis is a consequence
of `quote(a^(b + c))` actually producing a nested R\ call of the
form `'^'(a, (b + c))` instead of `'^'(a, b + c)`:

@@

This feature is unfortunate because the extra parenthesis around `b + c` is not
needed and yields to the extra parenthesis. The preferred result is obtained by
using the functional representation `quote('^'(k + l, m + n))` of the power, or,
as a workaround, curly braces (see last summand above).

@@

While the parentheses typically follow the precedence in standard infix
operators, undesirable results may be obtained in custom operators.

```{r, results='asis'}
term <- quote(mean(X) %+-% 1.96 * s / sqrt(N))
mathout(term)
```

This is a reminder that it is not possible to change the precedence of custom
operators in R\, and that expressions with such operators are evaluated strictly
from left to right. As in the previous example, the solution is to either use
the functional form of the operator or a curly brace to enforce the correct
operator precedence.

```{r, results='asis'}
term <- quote(`%+-%`(mean(X), 1.96 * s / sqrt(N)))
term <- quote(mean(X) %+-% {1.96 * s / sqrt(N)})   # the same
mathout(term)
```

More operators are shown in Table \@ref(tab:custom-operators), following the
suggestions by Murrell and Ihaka [-@murrell2000].

```{r custom-operators, echo=FALSE}
op1 <- list(
  cross=quote(x %*% y),
  dot=quote(x %.% y),
  kronecker=quote(X %x% Y),
  'integer division'=quote(x %/% y),
  and=quote(A & B),
  vel=quote(A | B),
  xor=quote(xor(A, B)),
  not=quote(!A),
  equal=quote(X == Y),
  assign=quote(X <- Y))

@@

op2 <- list(
  different=quote(x != y),
  tilde=quote(X ~ Y),
  approx=quote(x %~~% y),
  equiv=quote(x %==% y),
  congr=quote(x %=~% y),
  prop=quote(x %prop% y),
  'in'=quote(a %in% A),
  intersect=quote(intersect(A, B)),
  union=quote(union(A, B)),
  intersect=quote(is.null(A)))

@@

op3 <- list(
  twosidearrow=quote(x%<->%b),
  rightarrow=quote(x%->%y),
  leftarrow=quote(x%<-%y),
  uparrow=quote(x%up%y),
  down=quote(x%down%y),
  iff=quote(x%<=>%y),
  rArr=quote(x %=>% y),
  lArr=quote(x %<=% y),
  dblup=quote(x %dblup% y),
  dbldown=quote(x %dbldown% y))
m3 <- lapply(op3, FUN=mathout, flags=list(cat=FALSE))

@@

t <- cbind(Operator=op1, Output=m1, Operator=op2, Output=m2, Operator=op3, Arrow=m3)
knitr::kable(t, caption="Custom operators in mathml", 
             row.names=FALSE, escape=FALSE)
```

## R functions

All functions from R::base are supported, a subset is shown in 
Tables \@ref(tab:basic-elements).

```{r basic-elements, echo=FALSE}
op1 <- list(
  sin=quote(sin(x)),
  cosh=quote(cosh(x)),
  tanpi=quote(tanpi(alpha)),
  log=quote(log(p)),
  log1p=quote(log1p(x)),
  exp=quote(exp(x)),
  expm1=quote(expm1(x)),
  lchoose=quote(lchoose(n, k)),
  sqrt=quote(sqrt(x)),
  mean=quote(mean(xy)),
  absolute=quote(abs(x)))
m1 <- lapply(op1, FUN=mathout, flags=list(cat=FALSE))

@@

t <- cbind(Function=op1, Output=m1)
knitr::kable(t, caption="Arithmetic functions", row.names=FALSE, escape=FALSE)
```

@@

R\'s `integrate` function takes a number of arguments, the most important ones
being the function to integrate, and the lower and the upper bound of the
integration. For mathematical typesetting in the form
of $\int f(x)\, dx$, \pkg{mathml} needs to find out the name of the integration
variable. For that purpose, the underlying Prolog bridge provides a
predicate `r_eval/2` that calls R\ from Prolog. In the example below, this
predicate is used for a call of `formalArgs(args(sin))`, which returns the names
of the function arguments of `sin`, namely, `x`.

@@

@@

We can then easily feed mixtures of partially named and positional arguments in
unusual order into the renderer.

```{r, echo=TRUE, results='asis'}
g <- function(u)
{
  sin(u)
}

# 
term <- call("integrate", lower=-Inf, upper=Inf, g)
attributes(term)$name = quote(g)
mathout(canonical(term))

>>>>

---
title: "mathml: Translate R expressions to MathML and LaTeX/MathJax"
date: "2022-12-09"
abstract: >
  This R\ package translates R\ objects to suitable elements in MathML or LaTeX,
  thereby allowing for a pretty mathematical representation of R\ objects and
  functions in data analyses, scientific reports and interactive web content. In
  the RMarkdown document rendering language, R\ code and mathematical content
  already exist side-by-side. The present package allows to use the same
  R\ objects for both data analysis and typesetting in documents or web content.
  This tightens the link between the statistical analysis and its verbal
  description or symbolic representation, which is another step towards
  reproducible science. User-defined hooks allow to extend the package by
  mapping specific variables or functions to new MathML and LaTeX entities. A
  few working examples illustrate the use of the package in a scientific report
  and in on an interactive web page.
draft: true
author:  
  # see ?rjournal_article for more information
  - name: Matthias Gondan
    affiliation: University of Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/gondan-rochon/index.html.en
    orcid: 0000-0001-9974-0057
    email:  Matthias.Gondan-Rochon@uibk.ac.at
  - name: Irene Alfarone
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/alfarone/index.html.en
    email: Irene.Alfarone@uibk.ac.at
    orcid: 0000-0002-8409-8900
    affiliation: University of Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib
preamble:
- \usepackage{longtable}
- \usepackage{cancel}
---

@@

The R\ extension of the markdown language [@Xie2020] enables reproducible
statistical reports with nice typesetting in HTML, Microsoft Word, and Latex.
Since recently [@R, version 4.2], R\'s manual pages include support for
mathematical expressions [@Sarkar2022], which already is a big improvement.
However, rules for the mapping of built-in language elements to their
mathematical representation are still lacking. So far, R\ expressions such
as `pbinom(k, N, p)` are printed as they are; pretty mathematical formulae
such as \(P_\mathrm{Bi}(X \le k; N, p)\) require explicit Latex commands, that
is, `P_\mathrm{Bi}\left(X \le k; N, p\right)`. Except for minimalistic examples,
these commands are tedious to type in and their source code is hard to read.

The present R\ package defines a set of rules for the automatic translation of
R\ expressions to mathematical output in RMarkdown documents [@Xie2020] and
ShinyApp webpages [@Chang2022]. The translation is done by an embedded Prolog
interpreter that maps nested expressions recursively to MathML and
LaTeX/MathJax, respectively. User-defined hooks enable to extend the set of
rules, for example, to represent specific R\ elements by custom mathematical
signs.

The main feature of the package is that the same R\ expressions and equations
can be used for both mathematical typesetting and calculations. This saves time
and reduces mistakes, as will be illustrated below.

The paper is organized as follows. Section\ 2 presents the technical background
of the package, including the two main classes of Prolog rules for translating
R\ objects to mathematical expressions. Section\ 3 illustrates the main features
of the \pkg{mathml} package, potential issues and workarounds using examples
from the day-to-day perspective of a user. Sections\ 4 and\ 5 present two case
studies for the use of the package in a scientific report and a ShinyApp web
page. Section\ 6 concludes with a discussion and ideas for further development.
A complete list of the R\ elements covered by the package is found in the
Appendix.

@@

The translation of R\ expressions to mathematical output is achieved through a
Prolog interpreter. Prolog is a classical logic programming language with many
applications in expert systems, computer linguistics and symbolic artificial
intelligence. The main strength of Prolog is its concise representation of facts
and rules for the representation of knowledge and grammar, as well as its
efficient built-in search engine for closed world domains. As it is well-known,
R\ is a statistical programming language for data analysis and statistical
modeling which is widely used in academia and industry. Besides the core
library, a lot of packages have been developed for all kinds of statistical
problems, including statistics-based artificial intelligence tools such as
neural networks for machine learning and deep learning. Whereas Prolog is weak
in statistical computation, but strong in symbolic manipulation, the converse
may be said for the R\ language. The \pkg{rolog} package [@rolog] bridges this
gap by providing an interface to a SWI-Prolog distribution in R. The
communication between the two systems is mainly in the form of queries from
R\ to Prolog, but two predicates allow Prolog to ring back and evaluate terms in
R.

For a first illustration of the \pkg{mathml} package, we consider the binomial
probability.

@@

The term is quoted to avoid its immediate evaluation (which would raise an error
anyway since the variables `k`, `N`, `p` have not yet been defined). Experienced
readers will remember that the quoted expression above is a short form for

@@

As is seen from the output above, the variable `term` is not assigned the result
of the calculation, but an R\ "call"
[see, e.g., @Wickham2019, for details on "non-standard evaluation"]. This call
can eventually be evaluated with `eval()`,

@@

The R\ package \pkg{mathml} can now be used to render the call in MathML, that
is the dialect for mathematical elements on HTML webpages or in MathJax/LaTeX,
as shown below (some of the curly braces are not really needed in this simple
example, but are necessary in edge cases).

@@

We can include the output in a RMarkdown document by specifying `results='asis'`
in the R\ code chunk, as is done in the next example. The
R\ function `mathout()` is a wrapper that invokes `mathml()` for HTML output
and `mathjax()` for LaTeX output.

@@

At the Prolog end, a predicate `math/2` translates the call `pbinom(K, N, Pi)`
into a "function" `fn/2` with the name `P_Bi`, one argument `X =< K`, and the
two parameters `N` and `Pi`.

@@

Thus, the predicate `math/2` could be considered a "macro" that translates a
mathematical element (here, `pbinom(K, N, Pi)`) to a different mathematical
element, namely `fn(Name, (Args ; Pars))`. The low-level predicate `ml/3` is
used to convert these basic elements to MathML.

```prolog
ml(Flags, fn(Name, (Args ; Pars)), M)
 => ml(Flags, Name, N),
    ml(Flags, paren(list(op(;), [list(op(','), Args), list(op(','), Pars)])), X),
    M = mrow([N, mo(&(af)), X]).
```

The relevant rule for `ml/3` builds the MathML entity `mrow([N, mo(&(af)), X])`,
with `N` representing the name of the function and `X` its arguments and
parameters, enclosed in parentheses. A corresponding rule `jax/3` does the same
for MathJax/LaTeX. A list of flags can be used for context-sensitive
translation (see, e.g., the errors below).

@@

\pkg{mathml} handles the basic elements of everyday mathematical expressions,
such as numbers, Latin and Greek letters, multi-letter identifiers, accents,
subscripts, and superscripts.

```{r, results='asis'}
term <- quote(1 + -2L + a + abc + "a" + phi + Phi + varphi + hat(b)[i, j]^2L)
mathout(term)

@@

An expression such as `1 + -2` may be considered unsatisfactory from an
aesthetical perspective. It is correct R\ syntax, though, and is reproduced
accordingly, without the parentheses. Parentheses around negated numbers or
symbols can be added as shown for `-Inf`.

Note that an R function `hat()` does not exist in base R, it is provided by the
package for convenience and points to the identity function.

@@

Arithmetic operators and parentheses are translated as they are, as illustrated
below.

```{r, results="asis"}
term <- quote(a - ((b + c)) - d*e + f*(g + h) + i/j + k^(l + m) + (n*o)^{p + q})
mathout(term)

term <- quote(dot(d, e) + nodot(f, g + h) + times(X, Y))
mathout(term)
```

For multiplications involving only numbers and symbols, the multiplication sign
is omitted. This heuristic does not always produce the desired result;
therefore, \pkg{mathml} defines alternative multiplication
functions `dot()`, `nodot()`, and `times()` that produce the respective
multiplication signs.

For standard operators with known precedence, \pkg{mathml} is generally able to
detect if parentheses are needed; for example, parentheses are automatically
placed around `g + h` in the `nodot`-example. However, we note unecessary
parentheses around `l + m` above. Thes parentheses are a consequence
of `quote(a^(b + c))` actually producing a nested R\ call of the
form `'^'(a, (b + c))` instead of `'^'(a, b + c)`:

@@

For the present purpose, this feature is unfortunate because extra parentheses
around `b + c` are not needed. The preferred result is obtained by using the
functional form `quote('^'(k, l + m))` of the power, or curly braces as a
workaround (see `p + q` above).

@@

Whereas in standard infix operators, the parentheses typically follow the rules
for precedence, undesirable results may be obtained in custom operators.

```{r, results='asis'}
term <- quote(mean(X) %+-% 1.96 * s / sqrt(N))
mathout(term)
term <- quote('%+-%'(mean(X), 1.96 * s / sqrt(N)))
term <- quote(mean(X) %+-% {1.96 * s / sqrt(N)})   # the same
mathout(term)
```

The example is a reminder that it is not possible to define the precedence of
custom operators in R, and that expressions with such operators are evaluated
strictly from left to right. As in the previous example, the solution is to
either use the functional form of the operator or a curly brace to enforce the
correct operator precedence.

More operators are shown in Table \@ref(tab:custom-operators), including the
suggestions by Murrell and Ihaka [-@murrell2000] for graphical annotations and
arrows in R\ figures.

```{r custom-operators, echo=FALSE}
op1 <- list(
  cross=quote(A %*% B),
  dot=quote(A %.% B),
  kronecker=quote(A %x% B),
  'integer division'=quote(A %/% B),
  and=quote(A & B),
  vel=quote(A | B),
  xor=quote(xor(A, B)),
  not=quote(!A),
  equal=quote(A == B),
  assign=quote(A <- B))

@@

op2 <- list(
  different=quote(A != B),
  tilde=quote(A ~ B),
  approx=quote(A %~~% B),
  equiv=quote(A %==% B),
  congr=quote(A %=~% B),
  prop=quote(A %prop% B),
  'in'=quote(A %in% B),
  intersect=quote(intersect(A, B)),
  union=quote(union(A, B)),
  intersect=quote(is.null(A)))

@@

op3 <- list(
  twosidearrow=quote(A %<->% B),
  rightarrow=quote(A %->% B),
  leftarrow=quote(A %<-% B),
  uparrow=quote(A %up% B),
  down=quote(A %down% B),
  iff=quote(A %<=>% B),
  rArr=quote(A %=>% B),
  lArr=quote(A %<=% B),
  dblup=quote(A %dblup% B),
  dbldown=quote(A %dbldown% B))

m3 <- lapply(op3, FUN=mathout, flags=list(cat=FALSE))

@@

t <- cbind(Operator=op1, Output=m1,
  Operator=op2, Output=m2,
  Operator=op3, Arrow=m3)

knitr::kable(t, caption="Custom operators in mathml",
  row.names=FALSE, escape=FALSE)
```

## Builtin functions

There is support for most functions from \pkg{base}, with adequate use and omission
of parentheses.

```{r, results='asis'}
term <- quote(sin(x) + sin(x)^2L + cos(pi/2L) + tan(2L*pi) * expm1(x))
mathout(term)
term <- quote(choose(N, k) + abs(x) + sqrt(x) + floor(x) + ceiling(x) + round(X))
mathout(term)
```

A few more examples are shown in Table \@ref(tab:base-stats), including
functions from \pkg{stats}.

```{r base-stats, echo=FALSE}
op1 <- list(
  sin=quote(sin(x)),
  cosh=quote(cosh(x)),
  tanpi=quote(tanpi(alpha)),
  log=quote(log(p)),
  log1p=quote(log1p(x)),
  exp=quote(exp(x)),
  expm1=quote(expm1(x)),
  lchoose=quote(lchoose(n, k)),
  sqrt=quote(sqrt(x)),
  mean=quote(mean(X)["A"]),
  absolute=quote(abs(x)))
m1 <- lapply(op1, FUN=mathout, flags=list(cat=FALSE))

@@

op2 <- list(
  sin=quote(sin(x)),
  cosh=quote(cosh(x)),
  tanpi=quote(tanpi(alpha)),
  log=quote(log(p)),
  log1p=quote(log1p(x)),
  exp=quote(exp(x)),
  expm1=quote(expm1(x)),
  lchoose=quote(lchoose(n, k)),
  sqrt=quote(sqrt(x)),
  mean=quote(mean(X)["A"]),
  absolute=quote(abs(x)))
m2 <- lapply(op2, FUN=mathout, flags=list(cat=FALSE))

t <- cbind(Function=op1, Output=m1, Function=op2, Output=m2)
knitr::kable(t, caption="R functions from base and stats",
  row.names=FALSE, escape=FALSE)
```

## Custom functions

@@

R\'s `integrate` function takes a number of arguments, the most important ones
being the function to integrate, and the lower and the upper bound of the
integration.

@@

In the above example, `sin` is an R symbol, not an R function. For mathematical
typesetting in the form of $\int f(x)\, dx$, \pkg{mathml} needs to find out the
name of the integration variable. For that purpose, the underlying Prolog bridge
provides a predicate `r_eval/2` that calls R\ from Prolog. In the example above,
this predicate is used for a call of `formalArgs(args(sin))`, which returns the
names of the arguments of `sin`, namely, `x`.

If integrand points to an R function as in `call("integrate", sin, 0, 2L*pi)`,
mathml would be unable to find out the function name:

```{r}
g <- function(x) {sin(x)}
print(g)
```

@@

This function can be used to feed mixtures of partially named and positional
arguments into the renderer.

```{r, results='asis'}
g <- function(u)
{
  sin(u)
}

# 
term <- call("integrate", lower=-Inf, upper=Inf, g)
attributes(term)$name <- quote(g)
mathout(canonical(term))
